---
title: "Verified interval arithmetic in pure R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Verified interval arithmetic in pure R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(RobustArithmetic)
```

## What this package guarantees, and what it does not

Every operation in this package returns an **enclosure**: an interval that provably contains
the exact real result. Not an estimate of it, not an error bar around it. If the true answer
is a real number `y`, then `y` lies inside the returned interval, and that statement is a
theorem about the code rather than an observation about a sample of inputs.

The guarantee is deliberately narrow. An enclosure that contains the answer may still be
wide, and a wide enclosure is honest but not useful. Throughout the package the two
properties are kept apart and reported apart: **validity** is never traded for **tightness**,
and where a result cannot be tightened the package says so instead of narrowing it.

Conformance with IEEE Std 1788.1-2017 is **not** claimed, for the reason the standard itself
gives: its subclause 1.5 makes conformance a list of requirements an implementation shall
satisfy, with no partial grade available. What is followed, measured one requirement at a
time, is documented in `?RobustArithmetic` and summarised at the end of this vignette.

## The kernel: intervals and outward rounding

An interval is built from its endpoints and carries a decoration that records what is known
about the operation that produced it.

```{r}
x <- ra_interval(1, 2)
y <- ra_interval(3, 4)
ra_add(x, y)
ra_mul(x, y)
```

The endpoints of a computed interval are rounded **outward**, so the enclosure can only grow
and never silently shrink past the true result. R offers no way to change the processor's
rounding mode, and changing it would not be a local act in any case: it is per-thread state
of the processor, so it would govern every floating-point operation executed afterwards on
that thread, inside this package or anywhere else. Outward rounding is therefore obtained
arithmetically, from the predecessor and successor formulas of Rump, Zimmermann, Boldo and
Melquiond (2009), which are valid under round-to-nearest.

```{r}
ra_pred(1)
ra_succ(1)
```

Division by an interval containing zero is where a naive implementation quietly returns
nonsense. Here it returns the mathematically correct answer, which is a *pair* of intervals,
through the extended division of the interval Newton literature.

```{r}
ra_div_extended(ra_interval(1, 2), ra_interval(-1, 1))
```

## Elementary functions at two levels

Elementary functions are provided at two levels, and the difference between them is a
difference in what can be proved rather than a difference in speed alone.

The **fast** level evaluates over the system math library and widens the result by a declared
slack derived from published accuracy measurements. The **rigorous** level evaluates over
`Rmpfr` with a directed-rounding bridge. When a verdict would otherwise fall inside the slack
— that is, when the fast level cannot decide the question being asked — the package escalates
through a ladder of precisions rather than returning a confident answer it cannot defend.

```{r}
ra_elem("exp", ra_interval(0, 1))
ra_has_mpfr()
```

The slack is not a tuning knob chosen for convenience. It is measured against the library
actually installed, which matters more than it sounds: R does not call the system libm along
a single fixed path, so a bound published for one library does not certify what R computes.
`ra_measure_library_error()` is the instrument that measures it.

## Expressions, and why the centered form exists

Evaluating an expression by substituting an interval for each occurrence of the variable —
the **natural extension** — is valid but often loose, because each occurrence is treated as
if it varied independently of the others. The **centered** extension trades that looseness
for a derivative bound.

```{r}
ra_eval_natural(quote(x^2 - x), list(x = ra_interval(0, 1)))
ra_enclose_expr(quote(x^2 - x), ra_interval(0, 1))
```

The natural extension gives `[-1, 1]`; `ra_enclose_expr()` gives `[-0.75, 0.25]`, which is the
exact range. Neither form is uniformly better and the package does not pretend otherwise, so
`ra_enclose_expr()` takes the best of the methods it is given —- natural, mean value and a
monotonicity test -— rather than committing to one.

## Roots: the interval Newton operator

`ra_solve()` finds roots of a univariate expression over a box and returns, for each one, a
word about what was established. The engine is the Hansen-Sengupta operator with extended
division and epsilon-inflated candidate verification, driven by a subdivision (paving) loop.

```{r}
ra_solve(quote(x^2 - 2), ra_interval(0, 2))
```

The important property is the failure mode. The loop has a budget, and when the budget runs
out it **abstains by name** and prints what it spent, rather than returning a smaller set of
roots as though the search had been exhaustive. An abstention is information; a silently
truncated answer is not.

## Fixed points in higher dimensions, without interval matrices

The solver above is univariate by contract. For fixed points of maps on `R^n` the package
takes a different route, and the route is the point of the design: the whole certificate is a
**scalar inequality**, so a fixed point in `R^n` is certified without interval matrices,
without factorisations and without an interval spectral decomposition. The norm collapses the
dimension into three numbers, and the caller may bound those three numbers by any means they
can defend.

```{r}
ra_ball_certificate(rho = 0.2, lipschitz = 0.5)
```

Two theorems are applied and the object says which one decided the verdict. **Confinement**
— the map carries the ball into itself — gives existence by Brouwer and says nothing about
what happens inside. **Contraction** adds uniqueness and convergence by Banach. They are
certified separately because they are not the same statement, and a package that reported one
word for both would be reporting less than it knows.

```{r}
ra_ball_certificate(rho = 0.24, lipschitz = 1.6, radius = 0.4, image_radius = 0.4)
```

That second call is confined but not contractive: the image radius route needs no Lipschitz
constant, which is why it is not redundant.

Two assemblers help build the three inputs in the matrix case without factorising anything:
`ra_gershgorin()` encloses eigenvalues from the matrix entries, and `ra_spectral_sum()` bounds
a spectral quantity of a sum from bounds on its terms.

```{r}
ra_gershgorin(matrix(c(2, 0.1, 0.1, 3), 2, 2))
```

**The responsibility that cannot be delegated:** the three bounds must be measured in the same
norm, and the package cannot check that. The label travels with the object. Mixing norms
silently would be a failure mode indistinguishable, at run time, from a valid certificate.
Likewise the Lipschitz constant must hold over the whole ball and not merely at its centre; a
caller who bounds the derivative at the centre receives a certificate that means nothing, and
no arithmetic can detect it. In one dimension the package closes that loop itself, in
`ra_certify_fixed_point()`.

## A refusal certifies nothing

This is worth stating on its own, because it is the most common way a verified result is
misread. The hypotheses behind every certificate here are **sufficient and not necessary**.
There is no test of exclusion in this package and no word of demonstrated absence. A map that
this package declines to certify may still have a fixed point in the ball; a root that the
solver does not report may still exist inside a box the budget never reached. What the
package proves, it proves; what it declines, it declines without claiming the opposite.

## What is covered, and what is not

Measured against IEEE Std 1788.1-2017, one requirement at a time: the interval type and the
decoration system of clause 5; 22 of the 39 arithmetic operations of Table 4.1; and the seven
numeric functions of Table 4.3. What is **not** provided is the cancellative operations, the
interval comparison relations, the text input and output of subclause 6.8, the interchange
representation of subclause 7.3, and the tightest accuracy that subclause 6.5.2 requires of
the basic operations — which here are one unit in the last place wider at each end.

`ra_operator_table()` prints the coverage as data rather than as prose, and
`ra_verify_operator_table()` checks the claim against the code.

## References

Hansen, E., & Walster, G. W. (2004). *Global optimization using interval analysis* (2nd ed.).
Marcel Dekker.

Neumaier, A. (1990). *Interval methods for systems of equations*. Cambridge University Press.

Rump, S. M. (2010). Verification methods: Rigorous results using floating-point arithmetic.
*Acta Numerica*, 19, 287-449.

Rump, S. M., Zimmermann, P., Boldo, S., & Melquiond, G. (2009). Computing predecessor and
successor in rounding to nearest. *BIT Numerical Mathematics*, 49(2), 419-431.
